Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-08-05 review_article.Rmd

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document: default
fontsize: 11pt
link-citations: yes
fig_caption: yes
biblio-style: vancouver
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!--

@@

```{r ci-viz2, fig.cap = "Example of CI visualisation for uncertainty representation in cancer mapping (1/3). Source: Alberta Health IHDA Geographic. (2012) ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width='50%'}

@@

```{r ci-viz3, fig.cap = "Example of CI visualisation for uncertainty representation in cancer mapping (2/3). Source: Pensylvania Cancer Atlas ", echo=FALSE, message=FALSE, warning=FALSE, out.width='50%', fig.align='center'}

@@

```{r ci-tooltip, fig.cap= "Example of CI visualisation for uncertainty representation in cancer mapping (3/3). Source: Centres for Disease Control and Prevention (CDC). United States Cancer Statistics: An Interactive Cancer Atlas (InCA) ", echo=FALSE, message=FALSE, warning=FALSE, fig.show='hold', out.width= '50%', fig.align = 'center'}

@@

```{r ci-table, fig.cap= "Example of an interactive data table with CI upper and lower bounds used in cancer mapping. Source: Pennsylvania Cancer Atlas", echo=FALSE, message=FALSE, warning=FALSE, fig.show='hold', out.width= '50%', fig.align = 'center'}

@@

```{r boxplot-example, fig.cap= "Example of boxplot used in cancer mapping. Source: Source: Pensylvania Cancer Atlas ", echo=FALSE, message=FALSE, warning=FALSE, fig.show='hold', out.width= '40%', fig.align = 'center'}

@@

A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult [@AOCAFE].
The distortion of a cartogram prevents the density obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

@@

## Centroid displays

Dot plot: one dot for each region, coloured, and placed at centroid.

@NAC gives an example of

Plotting centroids on top of geographies. (Size is kept constant)

Replies on the idea that every area is important, no matter the size.
This gives equal emphasis to every area, allows distributions and relationships between neighbours to become more clear.

<!-- 
TVSSS parliamentary cartogram where each constituency is represented by an oval of equal size. Many ovals overlap because the cartogram is still in the process of being made. Each
-->

@@

There are two keys to drive the choice of display: the properties of the visualisation, and the ease or accuracy of information extraction for map users @SAMGIS.

<!-- @CPISACA contrasted several online atlases in a 2006 paper, consider for structure? -->

(only worth including if it is possible for us to implement)
Tabular form comparing and contrasting
- Relationship to geography

- Show using cancer examples

@@

Recent developments of technology allowed interactive

'Where control of the message is important, static maps will continue to be the most effective, although good tables, graphs, and explanatory text are still needed in order to ensure that different people will see the same thing in the maps' @CPISACA

lends to temporal pattern exploration

@ECGC ?? Highlight the value of animating contiguous to see changes over time, US can be recognisable but animation aides interpretation

>>>>

---
title: "Cartogram Mapping and its Application to Cancer Data Visualisation"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
fontsize: 11pt
link-citations: yes
fig_caption: yes
biblio-style: vancouver
yolo: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
<!--

@@

```{r ci-viz2, fig.cap = "Example of CI visualisation for uncertainty representation in cancer mapping (1/3). Source: Alberta Health IHDA Geographic. (2012) ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', out.width='50%', eval = FALSE}

@@

```{r ci-viz3, fig.cap = "Example of CI visualisation for uncertainty representation in cancer mapping (2/3). Source: Pensylvania Cancer Atlas ", echo=FALSE, message=FALSE, warning=FALSE, out.width='50%', fig.align='center', eval = FALSE}

@@

```{r ci-tooltip, fig.cap= "Example of CI visualisation for uncertainty representation in cancer mapping (3/3). Source: Centres for Disease Control and Prevention (CDC). United States Cancer Statistics: An Interactive Cancer Atlas (InCA) ", echo=FALSE, message=FALSE, warning=FALSE, fig.show='hold', out.width= '50%', fig.align = 'center', eval = FALSE}

@@

```{r ci-table, fig.cap= "Example of an interactive data table with CI upper and lower bounds used in cancer mapping. Source: Pennsylvania Cancer Atlas", echo=FALSE, message=FALSE, warning=FALSE, fig.show='hold', out.width= '50%', fig.align = 'center', eval = FALSE}

@@

```{r boxplot-example, fig.cap= "Example of boxplot used in cancer mapping. Source: Source: Pensylvania Cancer Atlas ", echo=FALSE, message=FALSE, warning=FALSE, fig.show='hold', out.width= '40%', fig.align = 'center', eval = FALSE}

@@

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualisation method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC].
Cartograms may be seen as an extension of map transformations and projections. The favourable distortion is proportional to a value other other than actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, making interpretation of spatial patterns very difficult.
The distortion of a cartogram prevents the density obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasises the data distribution instead of land size [@CBATCC].
When visualising population statistics @ACTUC considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
@CTTMB suggest that spatial socio-economic data is best presented on a cartogram for urban areas.
@HEDP agrees that 'cancer occurs in people, not in geographical areas' and the map bases of population reflect this and avoid allocating 'undue prominence' to rural areas. 
@MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities.

@@

@@

There are two keys presented by @SAMGIS to drive the choice of display: 
 - the properties of the visualisation, and 
 - the ease or accuracy of information extraction for map users

<!-- @CPISACA contrasted several online atlases in a 2006 paper, consider for structure?

(only worth including if it is possible for us to implement)
Tabular form comparing and contrasting
- Relationship to geography
- Show using cancer examples
-->

@@

Recent developments in technology allowed interactive web atlases.

> 'Where control of the message is important, static maps will continue to be the most effective, although good tables, graphs, and explanatory text are still needed in order to ensure that different people will see the same thing in the maps' @CPISACA

The intention of interactivity and animation is to allow users to answer their own questions.

<!--
Using e-Delphi to Evaluate the Pennsylvania Cancer Atlas - explains PCA and reasons for interactivity.
-->

<!--
@ECGC ?? Highlight the value of animating contiguous to see changes over time, US can be recognisable but animation aides interpretation 
-->
